---
title: "Phase1_submission"
author: "Emilie Wilkie"
date: "22/01/2020"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## data io
library(xlsx)

## plotting
library(ggplot2)
library(reshape)

## deg
library(limma)
library(DESeq)
library(edgeR)

## modelling
library(dplyr) # for data manipulation
#update.packages(oldPkgs="caret", ask=FALSE)
library(caret) # for model-building
library(DMwR) # for smote implementation
library(purrr) # for functional programming (map)
library(pROC) # for AUC calculations

```

# Subchallenge 1 - Gene Expression

## Cleaning data

### Annotation based filtering
Essentially need to subset the gene list. First extract all genes which are matched based on gene symbol int he HGNC database. This means that there will be other information available for these genes. 

```{r }
Phase1_GE <- read.table("/Users/ewilkie/Documents/PrecisionFDA/Data/sc1_Phase1_GE_FeatureMatrix.txt", sep="\t", header=T, row.names=1)
gene.names <- colnames(Phase1_GE)

## official symbols have - not .
gene.names_sub <- gsub("\\.", "-",gene.names)

## get data from symbol checker - generated from running the web tool (no API avail) on gene.names_sub
h <- read.xlsx("/Users/ewilkie/Documents/PrecisionFDA/Data/hgnc-symbol-check.xlsx", sheetName=1, header=T)

## find Unmatched
unmatched <- h[which(h$Match.type == "Unmatched"),]
print("Number of Unmatched gene names:")
dim(unmatched)
#[1] 719   7

## find Alias symbol
alias <- h[which(h$Match.type == "Alias symbol"),]

## find these can be duplicates of Approved symbol
dup <- h[which(h$Input %in% alias$Input),]
dup_name <- as.vector(dup$Input[duplicated(dup$Input)])

## these are not duplicated
aliasu <- alias[-which(alias$Input %in% dup_name),]

## combine to creat unique
hu <- h[-which(h$Input %in% alias$Input),]
hu <- rbind(hu, aliasu)
dim(hu)

```

### Plotting of expression values for filtered genes 
Does doing this exclude important information?

```{r }
unexp <- Phase1_GE[,which(colnames(Phase1_GE) %in% unmatched$Input)]
dim(unexp)

## order and colour data by mean
unexpo <- unexp[,order(apply(unexp, 2 , mean))]

## format for plottion
unexpm <- melt(cbind(rownames(unexpo),unexpo))
colnames(unexpm)[1] <- "Patient"

gbox <- ggplot(unexpm, aes(variable, value)) + geom_boxplot(position="dodge", outlier.size=.2) + ggtitle("Boxplot of Unmatched Genes with outliers for Gene Exprrssion in cohort") + theme(axis.text.x=element_blank())  + labs(x="Unmatched Gene", y="Expression")
print(gbox)
```

### Subset further - Extract only DEGs (p.value < 0.05)

```{r }
## formatting data
x <- as.matrix(Phase1_GE[,which(colnames(Phase1_GE) %in% hu$Input)])
eset <- ExpressionSet(assayData = t(x))

## for comparison
y <- read.table("/Users/ewilkie/Documents/PrecisionFDA/Data/sc1_Phase1_GE_Outcome.txt", sep="\t", header=T, row.names=1)
design <- model.matrix(~y$SURVIVAL_STATUS)

#data <- list(x,y)
#saveRDS(data, "sc1_Phase1_GE.rds")

## calculate
fit <- lmFit(eset, design)
fit <- eBayes(fit, trend=TRUE, robust=TRUE)
results <- decideTests(fit)
print(summary(results))

## extract results
top <- topTable(fit,n=Inf)
top.sig <- top[which(top$adj.P.Val < 0.05),]

print("Number of DEGs")
dim(top.sig)
##[1] 893   7

```

### PLot expression range of DEGs

```{r }
## boxplot of expresion values for only DEGs
sig <- x[,which(colnames(x) %in% rownames(top.sig))]

## format for plottion
unexpm <- melt(sig)
colnames(unexpm) <- c("Patient", "variable", "value")

## to get order by mean
unexpm$variable <- factor(unexpm$variable, ordered = FALSE )
unexpm$variable <- ordered(unexpm$variable, levels = colnames(sig)[order(apply(sig, 2 , mean))])

## plot data
gbox <- ggplot(unexpm, aes(variable, value)) + geom_boxplot(position="dodge", outlier.size=.2)  + ggtitle("Boxplot with outliers for Gene Exprrssion in cohort") + theme(axis.text.x=element_blank())  + labs(x="Gene", y="Expression")
print(gbox)
```

### Extract Properties from DEGs

Properties of interest are mean(expression) > 6 and SD > mean(sd(DEG)) for each gene


```{r }
## perform filtering
sub_by_mean <- x[,which(apply(x,2,mean) > 6)]
sub_by_sd <- sub_by_mean[,which(apply(sub_by_mean,2,sd) > mean(apply(sig,2,sd)))]
dim(sub_by_sd)

## save output
write.table(sub_by_sd,"/Users/ewilkie/Documents/PrecisionFDA/Data/Sc1_GE_filtered.csv",  sep=",")

## format for plottion
unexpm <- melt(sub_by_sd)
colnames(unexpm) <- c("Patient", "variable", "value")

## to get order by mean
unexpm$variable <- factor(unexpm$variable, ordered = FALSE )
unexpm$variable <- ordered(unexpm$variable, levels = colnames(sig)[order(apply(sig, 2 , mean))])

## plot data
gbox <- ggplot(unexpm, aes(variable, value)) + geom_boxplot(position="dodge", outlier.size=.2)  + ggtitle("Boxplot with outliers for Gene Expression in samples used for modelling") + theme(axis.text.x=element_blank())  + labs(x="Gene", y="Expression")
print(gbox)
```


## Data Modelling


### Build models and assess quality. Select one for downstream.
Create a model using three different methods. Stochastic Gradient Boosting (GMB), K-nearest neighbours (KNN) and Naive Bayes (NB). These methods were choose after examining Phenotype data for patients. Comparing against KNN and NB allows for assessment of overfitting. Typically if KNN AUC = 1 overfitting has occured. 

```{r }
## build classifier control
ctrl <- trainControl(method = "repeatedcv",number = 10,repeats = 5, summaryFunction = twoClassSummary, classProbs = TRUE)

## change outcome variable coding
y[which(y$SURVIVAL_STATUS == 0),] <- "Alive"
y[which(y$SURVIVAL_STATUS == 1),] <- "Dead"

## set operational parameters
set.seed(5627)
options(expressions = 5e5)

## format data
x_sub <- sub_by_sd
data <- merge(x_sub, y, by="row.names")
rownames(data) <- data[,1]
data <- data[,-1]
colnames(data)[ncol(data)] <- "Class"
dim(data)

## train data with gbm, knn and nb
gbm_fit <- train(Class ~ ., data = data, method = "gbm",verbose = FALSE, metric = "ROC", trControl = ctrl)
knn_fit <- train(Class ~ ., data = data, method = "knn", metric="ROC", trControl = ctrl)
nb_fit <- train(Class ~ ., data = data, method = "naive_bayes", metric="ROC", trControl = ctrl)
glmnet_fit <- train(Class ~ ., data = data, method = "glmnet", metric="ROC", trControl = ctrl)
```

### Plot AUC for model evelation and selection

```{r }
## extract and combine relevant data
roc_data_gbm <- roc(data$Class,predict(gbm_fit, data, type = "prob")[,"Dead"])
pd_gbm <- cbind(roc_data_gbm$specificities,roc_data_gbm$sensitivities, "GBM")
colnames(pd_gbm) <- c("specificities","sensitivities", "method")

roc_data_knn <- roc(data$Class,predict(knn_fit, data, type = "prob")[,"Dead"])
pd_knn <- cbind(roc_data_knn$specificities,roc_data_knn$sensitivities, "KNN")
colnames(pd_knn) <- c("specificities","sensitivities", "method")

roc_data_nb <- roc(data$Class,predict(nb_fit, data, type = "prob")[,"Dead"])
pd_nb <- cbind(roc_data_nb$specificities,roc_data_nb$sensitivities, "NB")
colnames(pd_nb) <- c("specificities","sensitivities", "method")

roc_data_glmnet <- roc(data$Class,predict(glmnet_fit, data, type = "prob")[,"Dead"])
pd_glmnet <- cbind(roc_data_glmnet$specificities,roc_data_glmnet$sensitivities, "glmnet")
colnames(pd_glmnet) <- c("specificities","sensitivities", "method")

pdfs <- rbind(pd_gbm,pd_knn,pd_nb,pd_glmnet)
pdfs <- as.data.frame(pdfs)

## format
pdfs$specificities <- as.numeric(as.vector(pdfs$specificities))
pdfs$sensitivities <- as.numeric(as.vector(pdfs$sensitivities))

## TO DO annotate with AUC value
ggplot(pdfs, aes(x=specificities,y=sensitivities,colour=method)) + geom_point(size=.5) + geom_line(size=1.15) + scale_x_continuous(trans = "reverse") + ggtitle("AUC - Comparisons on Stochastic Gradient Boosting (GBM), K-nearest neighbours (KNN) and Naive Bayes (NB)")
```


a) Provide a description of model settings and parameters, and details of model building including dataset(s) description(s) used for training, cross validation and testing (number of samples, number of features, etc.) 


## Detailed Best Model Accuracy Evaluation

```{r }
## predict data
pred <- predict(gbm_fit, data, type = "prob")

## find which ones do and don't match
comb <- cbind(pred, data$Class)

check <- list()
for (i in 1:nrow(comb)){
  ind <- which.max(comb[i,1:2])
  check[[i]] <- names(ind)
}

check_vec <- unlist(check)
merg <- cbind(comb, check_vec)

## classes don't match
wrong_class <- merg[which(merg[,3] != merg[,4]),]

##b) Short listed features selected by the model
impgenes <- summary(gbm_fit)[which(summary(gbm_fit)$rel.inf != 0),]
dim(impgenes)

## d) Confusion matrix indicating number of predictions, true positives, false positives, true negatives, and false negatives                       
Confusion.matrix <- table(y$SURVIVAL_STATUS,check_vec)
rownames(Confusion.matrix) <- c("TP", "FN")
colnames(Confusion.matrix) <- c("FP" ,"TN")
print(Confusion.matrix)

## e) Overall accuracy
error_rate <- nrow(wrong_class) / length(y$SURVIVAL_STATUS)
OA <- 1- error_rate
print(OA)

## f) Specificity g) Sensitivity 
w <- which(gbm_fit$results$n.trees == gbm_fit$bestTune$n.trees & gbm_fit$results$interaction.depth == gbm_fit$bestTune$interaction.depth & gbm_fit$results$shrinkage == gbm_fit$bestTune$shrinkage & gbm_fit$results$n.minobsinnode == gbm_fit$bestTune$n.minobsinnode)
ss <- gbm_fit$results[w,c("ROC","Sens","Spec")]
print(ss)

## h) Area under the curve (AUC)
## plot ROC with AUC label
plot(roc(data$Class,predict(gbm_fit, data, type = "prob")[,"Dead"]), print.auc=TRUE)
```
