---
title: "Phase1_submission"
author: "Emilie Wilkie"
date: "22/01/2020"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## plotting
library(ggplot2)
library(reshape)

## deg
library(limma)
library(DESeq)
library(edgeR)

## modelling
library(dplyr) # for data manipulation
#update.packages(oldPkgs="caret", ask=FALSE)
library(caret) # for model-building
library(DMwR) # for smote implementation
library(purrr) # for functional programming (map)
library(pROC) # for AUC calculations

```

# Subchallenge 3 - Gene Expression and CN

## Upload and format data

```{r }
## read in patient data
Phase1_CN_GE_out <- read.table("/Users/ewilkie/Documents/PrecisionFDA/Data/sc3_Phase1_CN_GE_FeatureMatrix.txt", sep="\t", header=T, row.names=1)
dim(Phase1_CN_GE_out)

y <- read.table("/Users/ewilkie/Documents/PrecisionFDA/Data/sc3_Phase1_CN_GE_Outcome.txt", sep="\t", header=T, row.names=1)

cyto.ind <- grep("^X[0-9]", colnames(Phase1_CN_GE_out))
cyto <- Phase1_CN_GE_out[,cyto.ind]
dim(cyto)
exp <- Phase1_CN_GE_out[,-cyto.ind]
dim(exp)
colnames(exp) <- gsub("\\.", "-",colnames(exp))

hu <- read.table("/Users/ewilkie/Documents/PrecisionFDA/Data/Sc1_GE_names.csv", sep=",")
exp_filt <- exp[,which(colnames(exp) %in% hu$Input)]

```

## Extract only DEGs (p.value < 0.05)

```{r }
## formatting data
x <- exp_filt
eset <- ExpressionSet(assayData = t(x))

design <- model.matrix(~y$SURVIVAL_STATUS)

## calculate
fit <- lmFit(eset, design)
fit <- eBayes(fit, trend=TRUE, robust=TRUE)
results <- decideTests(fit)
print(summary(results))

## extract results
top <- topTable(fit,n=Inf)
top.sig <- top[which(top$adj.P.Val < 0.05),]

print("Number of DEGs")
dim(top.sig)

```

### Extract Properties from DEGs

Properties of interest are mean(expression) > 6 and SD > mean(sd(DEG)) for each gene

```{r }
## perform filtering
sig <-  x[,which(colnames(x) %in% rownames(top.sig))]
sub_by_mean <- x[,which(apply(x,2,mean) > 6)]
sub_by_sd <- sub_by_mean[,which(apply(sub_by_mean,2,sd) > mean(apply(sig,2,sd)))]
dim(sub_by_sd)
```

## Format CN data

```{r }
x <- cyto
for(col in 1:ncol(x)){
  minv <- min(x[-which(x[,col] == 0),col], na.rm=T)
  x[which(x[,col] == 0),col] <- minv
}

x[x == NA] <- min(x)
x[x == Inf] <- min(x)

## check for SD
sd <- apply(x, 2 , sd)
x <- x[,-which(sd < 0.1)]

data <- cbind(sub_by_sd, x)
dim(data)
```


## Data Modelling


### Build models and assess quality. Select one for downstream.
Create a model using three different methods. Stochastic Gradient Boosting (GMB), K-nearest neighbours (KNN) and Naive Bayes (NB). These methods were choose after examining Phenotype data for patients. Comparing against KNN and NB allows for assessment of overfitting. Typically if KNN AUC = 1 overfitting has occured. 

```{r }
## build classifier control
ctrl <- trainControl(method = "repeatedcv",number = 10,repeats = 5, summaryFunction = twoClassSummary, classProbs = TRUE)

## change outcome variable coding
y[which(y$SURVIVAL_STATUS == 0),] <- "Alive"
y[which(y$SURVIVAL_STATUS == 1),] <- "Dead"

## set operational parameters
set.seed(5627)
options(expressions = 5e5)

## format data
x_sub <- data
data <- merge(x_sub, y, by="row.names")
rownames(data) <- data[,1]
data <- data[,-1]
colnames(data)[ncol(data)] <- "Class"
dim(data)

## train data with gbm, knn and nb
gbm_fit <- train(Class ~ ., data = data, method = "gbm",verbose = FALSE, metric = "ROC", trControl = ctrl)
knn_fit <- train(Class ~ ., data = data, method = "knn", metric="ROC", trControl = ctrl)
nb_fit <- train(Class ~ ., data = data, method = "naive_bayes", metric="ROC", trControl = ctrl)
glmnet_fit <- train(Class ~ ., data = data, method = "glmnet", metric="ROC", trControl = ctrl)
```

### Plot AUC for model evelation and selection

```{r }
## extract and combine relevant data
roc_data_gbm <- roc(data$Class,predict(gbm_fit, data, type = "prob")[,"Dead"])
pd_gbm <- cbind(roc_data_gbm$specificities,roc_data_gbm$sensitivities, "GBM")
colnames(pd_gbm) <- c("specificities","sensitivities", "method")

roc_data_knn <- roc(data$Class,predict(knn_fit, data, type = "prob")[,"Dead"])
pd_knn <- cbind(roc_data_knn$specificities,roc_data_knn$sensitivities, "KNN")
colnames(pd_knn) <- c("specificities","sensitivities", "method")

roc_data_nb <- roc(data$Class,predict(nb_fit, data, type = "prob")[,"Dead"])
pd_nb <- cbind(roc_data_nb$specificities,roc_data_nb$sensitivities, "NB")
colnames(pd_nb) <- c("specificities","sensitivities", "method")

roc_data_glmnet <- roc(data$Class,predict(glmnet_fit, data, type = "prob")[,"Dead"])
pd_glmnet <- cbind(roc_data_glmnet$specificities,roc_data_glmnet$sensitivities, "glmnet")
colnames(pd_glmnet) <- c("specificities","sensitivities", "method")

pdfs <- rbind(pd_gbm,pd_knn,pd_nb,pd_glmnet)
pdfs <- as.data.frame(pdfs)

## format
pdfs$specificities <- as.numeric(as.vector(pdfs$specificities))
pdfs$sensitivities <- as.numeric(as.vector(pdfs$sensitivities))

## TO DO annotate with AUC value
ggplot(pdfs, aes(x=specificities,y=sensitivities,colour=method)) + geom_point(size=.5) + geom_line(size=1.15) + scale_x_continuous(trans = "reverse") + ggtitle("AUC - Comparisons on Stochastic Gradient Boosting (GBM), K-nearest neighbours (KNN) and Naive Bayes (NB)")
```


a) Provide a description of model settings and parameters, and details of model building including dataset(s) description(s) used for training, cross validation and testing (number of samples, number of features, etc.) 


## Detailed Best Model Accuracy Evaluation

```{r }
## predict data
pred <- predict(gbm_fit, data, type = "prob")

## find which ones do and don't match
comb <- cbind(pred, data$Class)

check <- list()
for (i in 1:nrow(comb)){
  ind <- which.max(comb[i,1:2])
  check[[i]] <- names(ind)
}

check_vec <- unlist(check)
merg <- cbind(comb, check_vec)

## classes don't match
wrong_class <- merg[which(merg[,3] != merg[,4]),]

##b) Short listed features selected by the model
imp <- summary(gbm_fit)[which(summary(gbm_fit)$rel.inf != 0),]
write.table(imp, "/Users/ewilkie/Documents/PrecisionFDA/Data/Sc3_GE_CN_important_features.csv",  sep=",")

## d) Confusion matrix
Confusion.matrix <- confusionMatrix(as.factor(check_vec), as.factor(data$Class))
print(Confusion.matrix$table)

## e) Overall accuracy
error_rate <- nrow(wrong_class) / length(y$SURVIVAL_STATUS)
OA <- 1- error_rate
print(OA)

## f) Specificity 
spec <- Confusion.matrix$byClass[2]
print(spec)

#g) Sensitivity 
sens <- Confusion.matrix$byClass[1]
print(sens)

## h) Area under the curve (AUC)
## plot ROC with AUC label
plot(roc(data$Class,predict(gbm_fit, data, type = "prob")[,"Dead"]), print.auc=TRUE)
```
